Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-20 demystify.tex
updates

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs) and research ability, e.g., publications, statement of purpose (SOP).
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Publications in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipments and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective and have helped many students getting into top PhD programs in the US.

@@

\begin{commentbox}[Vu:]
  I often read LORs and SOP first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\paragraph{Publications} A concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has been successfully involved in research.  Applicants admitted to top schools often have multiple first-authored papers at top places.

@@

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, a 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is making these issues known in your SOP.  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Universities with Vietnamese profs} are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

\paragraph{Online Courses and Certificates}
In general, these do not carry much weight as they do not show research ability. We do not care much if you have taken on online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you might be able to use these to show you have sufficient CS knowledge.

@@

\textbf{Advice}: Prepare your CV in such away that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (busy) schedule. Do not be surprised if you get an interview invitation in the last minute. Some schools do not do interviews at all.

@@

\paragraph{Updating your profile} You should not send emails to update your profile.  However, if you have new publications or other big achievements, you can ask to update your application (though no guarantee that they will consider them).

@@

However, no interviews typically means that you will not likely get an RA (\autoref{sec:ra}), which is offered by an individual faculty (if they you to do research for them, then they likely will interview you first).  In the case of no interviews, your application (and TA/fellowship funding) is decided by the adcom.

@@

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.
\textbf{Advice}: Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, and talk to individual faculty and current students.  Take notes of faculty who make you excited, and count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, and funding situation.  Ask about anything you want to determine that they deserve \emph{you}.

@@

You can consider applying to MS programs, which are typically easier to get in; but you likely need to pay.  Sometimes, the university that rejected you for PhD might accept you for MS.
If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

\paragraph{Low Stipend?} Students often complain their stipend being too low. However, in most cases it is actually not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few would have NSF Graduate (GRFP) or Microsoft fellowships . In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, school will still mention you on their website, and you should put it on your CV.

@@

Most US universities have CS study, but many of them do not have a PhD program in CS. These universities might offer just Bachelor's degrees (e.g., BS) and no graduate studies (i.e., no MS or PhD degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has PhD in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, and Arlington, have PhDs in CS.

@@

\begin{commentbox}[Dat:] Most Vietnamese students, including those from top schools, \textbf{do not know} about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

While many find CSRankings useful, it is not perfect, and in general, rankings are superficial, and you need to do more research to be informed and make better decisions. For example, you might prefer places with a large community from your country (e.g., Northern VA has a large population of Vietnamese). You might prefer places with high-tech companies (Seattle or Silicon Valley), with many outdoor activities (e.g., hiking, skiing), or with better weather (\emph{``PhD can be depressing, so would you rather be depressed in California or New York?''}).  You might need to think about the cost of living, e.g., living in California is more expensive than in Virginia, or places with high crime rates (note that while some university might be in a high crime city, but the campus itself is safe).

@@

\begin{commentbox}[Xiaokuan:]
  Chinese students often only look at US News rankings when selecting their PhD universities (I did that, too, when I was applying for PhD positions).
  Now that I am a professor, I find it to be the least promising way.
  %
  The reason is that US News does not provide a good metric for evaluating the quality of the PhD program.
  %
  If you want to do great research, CSRankings is the best way to find good and active professors (which did not exist when I was applying),
  since it solely focuses on publications at top-tier CS conferences.
  %
  Also,
  I think PhD is not only about research;
  you need to also consider your daily life there, since you will (probably) stay for at least five years.
  %
  You might regret it if you did not consider this seriously before applying.
  %
\end{commentbox}

\section{Choosing an Adviser}
There is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with.  But how do you find such a person?

@@

Another effective way is taking graduate-level courses in the topics you are interested in.  Profs teaching \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof after classes, etc---being stand out.  Many profs, including myself, prefer taking in new students this way.  It gives both the prof and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).

@@

Not every faculty can serve as your official PhD adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these terminologies.

@@

\textbf{Tenure-line faculty}, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring PhD students.  They often have a low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  For most schools, tenure-line is the main type of faculty that can serve as a formal adviser of PhD students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

\textbf{Teaching faculty} (sometimes called instructional faculty or professor of practice) mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. As such they typically do not have funding (e.g., RA~\autoref{sec:funding}) to support PhD students.
Depends on the university, they might be able to \emph{co-advise} and mentor PhD students with a tenure-line faculty, but typically do not formally advise and graduate PhD students.

@@

Some universities have \textbf{research faculty} or research scientists, who are typically non-tenure line faculty and focus on research (e.g., publishing papers and getting grants). 
They can recruit students and provide RA funding for them. 
They can also teach classes, but typically have a lower teaching load.
For some universities, they \emph{can} solely advise and graduate PhD students, but for others, they cannot.

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with assistant, then promoted to associated, and full prof.

@@

Faculty often receive ``cold'' e-mails from prospective students. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that gets our attention?

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \autoref{sec:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and ask them if your paper has a chance.

So how to get someone to look at your profile and give input? You could ask your professors, LoR writers, collaborators, or those who have previously applied. For this kind of feedback, ask someone you have personal connection with.

@@

First, read the prof's website, see if they say something about contacting them. Many profs. explicitly indicate how prospective students should (or should not) contact them (e.g., using specific email subjects).
In general, the best way to catch the prof.'s attention is to \emph{customize your email} for them.  For example, read their papers, know what they work on, and see if you are interested in their research. Then email them and talk about how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email or copy and paste paper titles), you will not get a response.

@@

  This is a good example because it is clearly written just for me.  It shows that the student knows about my work on invariant generation and has related  background (paper C and project G).
\end{commentbox}

>>>>

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and ability, e.g., publications.
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Publications in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective and have helped many students get into top PhD programs in the US.

@@

\begin{commentbox}[Vu:]
  I often read LORs and SOP first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really stands out in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\paragraph{Publications} Concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has been successfully involved in research.  Applicants admitted to top schools often have multiple first-authored papers at top places.

@@

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, 3.5 GPA students from such talented programs are typically much better than 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is to make these issues known in your SOP.  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Universities with Vietnamese profs} are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

\paragraph{Online Courses and Certificates}
In general, these do not carry much weight as they do not show research ability. We do not care much if you have taken an online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you might be able to use these to show you have sufficient CS knowledge.

@@

\textbf{Advice}: Prepare your CV in such a way that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty set up interviews based on their (busy) schedule. Do not be surprised if you get an interview invitation at the last minute. Some schools do not do interviews at all.

@@

\paragraph{Updating your profile} You should not send emails to update your profile.  However, if you have new publications or other big achievements, you can ask them to update your application (though no guarantee that they will consider them).

@@

However, no interviews typically mean that you will not likely get an RA (\autoref{sec:ra}), which is offered by an individual faculty (if they want you to do research for them, then they likely will interview you first).  In the case of no interviews, your application (and TA/fellowship funding) is decided by the adcom.

@@

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.
\textbf{Advice}: Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, and talk to individual faculty and current students.  Take notes of faculty who make you excited, and count those taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, and the funding situation.  Ask about anything you want to determine that they deserve \emph{you}.

@@

You can consider applying to MS programs, which are typically easier to get in; but you likely need to pay.  Sometimes, the university that rejected you for PhD might accept you for MS.
If you get into an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

\paragraph{Low Stipend?} Students often complain their stipend is too low. However, in most cases it is not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about the stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few would have NSF Graduate (GRFP) or Microsoft fellowships. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, you school will still mention you on their website, and you should put it on your CV.

@@

Most US universities have CS studies, but many of them do not have a PhD program in CS. These universities might offer just Bachelor's degrees (e.g., BS) and no graduate studies (i.e., no MS or PhD degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has PhD in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiple locations of the University of Texas, e.g., Austin, Dallas, and Arlington, have PhDs in CS.

@@

\begin{commentbox}[Dat:] Most Vietnamese students, including those from top schools, \textbf{do not know} about CSRankings.  Maybe applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

While many find CSRankings useful, it is not perfect, and in general, rankings are superficial, and you need to do more research to be informed and make better decisions. For example, you might prefer places with a large community from your country (e.g., Northern VA has a large population of Vietnamese). You might prefer places with high-tech companies (Seattle or Silicon Valley), with many outdoor activities (e.g., hiking, skiing), or with better weather (\emph{``PhD can be depressing, so would you rather be depressed in California or New York?''}).  You might need to think about the cost of living, e.g., living in California is more expensive than in Virginia, or places with high crime rates (note that while some universities might be in a high-crime city but the campus itself is safe).

@@

\begin{commentbox}[Xiaokuan:]
  Chinese students often only look at US News rankings when selecting their PhD universities (I did that, too, when I was applying for PhD positions).
  Now that I am a professor, I find it to be the least promising way.
  %
  The reason is that US News does not provide a good metric for evaluating the quality of the PhD program.
  %
  If you want to do great research, CSRankings is the best way to find good and active professors (which did not exist when I was applying),
  since it solely focuses on publications at top-tier CS conferences.
  %
  Also,
  I think PhD is not only about research;
  you need to also consider your daily life there since you will (probably) stay for at least five years.
  %
  You might regret it if you did not consider this seriously before applying.
  %
\end{commentbox}

\section{Choosing an Adviser}
There is no one-size-fits-all answer to this question. The best adviser is the one that you can work well with.  But how do you find such a person?

@@

Another effective way is taking graduate-level courses in the topics you are interested in.  Profs teaching \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof. after classes, etc---being stand out.  Many profs, including myself, prefer taking in new students this way.  It gives both the prof and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).

@@

Not every faculty can serve as your official PhD adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, and research professors. Here is a primer on these terminologies.

@@

\textbf{Tenure-line faculty}, consisting of tenured and tenure-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring PhD students.  They often have a low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to permanent positions (informally, very hard to fire them).  \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  For most schools, tenure-line is the main type of faculty that can serve as a formal adviser of PhD students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

\textbf{Teaching faculty} (sometimes called instructional faculty or professor of practice) mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. As such they typically do not have funding (e.g., RA~\autoref{sec:funding}) to support PhD students.
Depending on the university, they might be able to \emph{co-advise} and mentor PhD students with a tenure-line faculty but typically do not formally advise and graduate PhD students.

@@

Some universities have \textbf{research faculty} or research scientists, who are typically non-tenure line faculty and focus on research (e.g., publishing papers and getting grants). 
They can recruit students and provide RA funding for them. 
They can also teach classes but typically have a lower teaching load.
For some universities, they \emph{can} solely advise and graduate PhD students, but for others, they cannot.

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with assistant, then is promoted to associated, and full prof.

@@

Faculty often receive ``cold'' e-mails from prospective students. Most of the time, we ignore these emails, but on some rare occasions, we do answer them. So how to write an email that gets our attention?

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \autoref{sec:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and asking them if your paper has a chance.

So how to get someone to look at your profile and give input? You could ask your professors, LoR writers, collaborators, or those who have previously applied. For this kind of feedback, ask someone you have a personal connection with.

@@

First, read the prof's website, and see if they say something about contacting them. Many profs. explicitly indicate how prospective students should (or should not) contact them (e.g., using specific email subjects).
In general, the best way to catch the prof.'s attention is to \emph{customize your email} for them.  For example, read their papers, know what they work on, and see if you are interested in their research. Then email them and talk about how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email or copy and paste paper titles), you will not get a response.

@@

  This is a good example because it is written just for me.  It shows that the student knows about my work on invariant generation and has a related background (paper C and project G).
\end{commentbox}
